Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data

### 2022-05-19 analysis/paper_draft2.Rmd
notes/questions re: revisions in draft text

@@

To evaluate the performance of the proposed method, we conducted a simulation study. The primary aim of these simulations is to examine the effect of the strength of discounting of the historical data by the power prior ($\alpha$), and the size of the historical data set, on our ability to detect bias.

@@

Figure 1 depicts the bias detection rate with an 95\% credible interval for the n_strikes = 15 scenario. When $b_{curr}$ is high (close to 3) the coverage rate is high, very close to 1. However, when $b_{curr}$ is lower (close to zero), the bias detection rate drops below 0.8. When there is no bias in the current or historical trials ($b_{curr} = b_{hist} = 0$}), our bias detection rate for 95% CI is 0.05 or less. As we place more weight on the historical strike data (as $a$ increases), and the historical strike data is consistent with current information (high compatibility between current and historical trials), the bias detection rate is close to 1, especially when bias is high (upper right-hand corner of plots). Conversely, as $\alpha$ decreases, bias detection is only high when there is high bias in the current trial ($b_{curr} \geq 1.5$). Increasing the number of historical trials leads to an increase in bias detection especially when current and historical trials are compatible.

@@

Here, most of the prosecutor's density curve of both prior and posterior are to the left of zero, indicating the possible bias against male jurors. However, because the 80\% credible interval includes 0, inferring such bias from the strike data alone is unjustified.

@@

Furthermore, the bias detection rate depends on the distance between the bias parameters for the current trial and for all the past (historical) trials ($ = |b_{curr} - b_{hist}|$), as our simulations show. These bias parameters may be incompatible ($b_{curr} \neq b_{hist}$) for at least two reasons.

@@

Second, even if that past strike data is complete (no missingness), incompatibility may occur if attorney bias depends on certain trial-level characteristics, such as defendant race or the severity of the criminal offenses charged, that take on one value in the current trial but that vary across the historical trials. For example, suppose an attorney who is more likely to strike Black prospective jurors because the defendant is Black; who has zero strike bias in cases with white defendants; and whose historical strike data comes from past trials with an equal number of Black and white defendants. If so, then bias detection in the current trial will depend on whether we use all the historical data or only the subset of past trials with Black defendants.

The simulations further show that bias detection increases with the number of an attorney's strikes per trial or the number of historical trials. This is in part because of our conservative initial prior for the bias parameter ($b \sim N(0,2)$) to accord with the legal burden of proving bias on those who challenge an attorney for strike bias. This accords with @Kadane2021, who recommends a neutral and symmetrical prior for estimating bias in _Batson_ challenges [FIX].  Thus, given little historical strike data for an attorney and a low number of strikes per trial, the initial prior has more sway and, a a result, the model we use will only detect severe bias.

END PARAGRAPH
generated a model that can identify bias and incorporates historical trial information - tested this with simulation
made an app to make the use of the model more accessible - hope is that lawyers will use it in practice

TODO: Add piece about small sample effect and fact that intial prior will dominate in small sample setting.

>>>>

<!-- IS THE ABOVE PARAGRAPH STILL ACCURATE?-->

@@

To evaluate the performance of the proposed method, we conducted a simulation study using Stan `r rstan::stan_version()`, a Hamiltonian Monte Carlo engine for Bayesian inference by way of `r R.version$version.string` and RStan [@rstan]. These simulations primarily show how much our ability to detect bias was affected by how much the power prior ($\alpha$) discounted the weight of the historical data and the size of the historical data.

@@

Figure 1 depicts the bias detection rate with an 95\% credible interval for the 15 strikes scenario. When $b_{curr}$ is high (close to 3) the coverage rate is high, very close to 1. However, when $b_{curr}$ is lower (close to zero), the bias detection rate drops below 0.8. When there is no bias in the current or historical trials ($b_{curr} = b_{hist} = 0$}), our bias detection rate for 95% CI is 0.05 or less. As we place more weight on the historical strike data (as $a$ increases), and the historical strike data is consistent with current information (high compatibility between current and historical trials), the bias detection rate is close to 1, especially when bias is high (upper right-hand corner of plots). Conversely, as $\alpha$ decreases, bias detection is only high when there is high bias in the current trial ($b_{curr} \geq 1.5$). Increasing the number of historical trials leads to an increase in bias detection especially when current and historical trials are compatible.

@@

Here, most of the prosecutor's density curve of both prior and posterior are to the left of zero, indicating the possible bias against male jurors. However, because the 80\% credible interval includes zero, inferring such bias from the strike data alone is unjustified.

@@

Furthermore, the bias detection rate depends on the distance between the bias parameters for the current trial and for all the past (historical) trials ($ = |b_{curr} - b_{hist}|$). As we explained above when we let $b_{hist}$ differ from $b_{curr}$ (our parameter of interest) in our simulations, these bias parameters may be incompatible ($b_{curr} \neq b_{hist}$) for at least two reasons.

@@

Second, even if that past strike data is complete (no missingness), incompatibility may occur if attorney bias depends on certain trial-level characteristics, such as defendant race or the severity of the criminal offenses charged, that take on one value in the current trial but that vary across the historical trials. For example, suppose an attorney who is more likely to strike Black prospective jurors because the defendant is Black; who has zero strike bias in cases with white defendants; and whose historical strike data comes from past trials with an equal number of Black and white defendants. If so, then bias detection in the current trial will depend on whether we use all the historical data or only the subset of past trials with Black defendants

The simulations further show that bias detection increases with the number of an attorney's strikes per trial or the number of historical trials. This is in part because of our conservative initial prior for the bias parameter ($b \sim N(0,2)$) to accord with the legal burden of proving bias on those who challenge an attorney for strike bias. This accords with @Kadane2021, who recommends a neutral and symmetrical prior for estimating bias in _Batson_ challenges [FIX].

<!-- what about Kadane's recommendation to use multiple priors for sensitivity? -->

Thus, given little historical strike data for an attorney and a low number of strikes per trial, the initial prior will dominate, and,  as a result, the model we use will only detect severe bias.

# Conclusion

generated a model that can identify bias and incorporates historical trial information - tested this with simulation
made an app to make the use of the model more accessible - hope is that lawyers will use it in practice

TODO: Add piece about small sample effect and fact that intial prior will dominate in small sample setting. <!-- WHAT to ADD to LAST paragraph in DISCUSSION section? -->
